Place of death in NRS COVID-19 deaths

### 2021-02-19 covid_nrs_place_of_death.Rmd
added small changes to home deaths all scotland plot to improve it; also added computation of annual ratios of all deaths to past deaths; started looking at number of covid cases by region and splitting regions by cases

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
```

@@

@@

```{r}
## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All") %>%
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets. The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other. The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other. This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
```

@@

```{r}
weekly_covid_cases_hb <-
  cases_hb %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )

weekly_covid_cases_la <-
  cases_la %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )
```

@@

# Calculate data averaged over multiple weeks

@@

@@

## Set up output directory

@@

```{r}
date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")

@@

## Deaths at home, all of Scotland

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
figure_deaths_at_home <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", size = 0.2) +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 12, height = 9, dpi = 300)
)

@@

We can use the yearly cumulative case count to group geographic areas by how many cases they've had

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
```

@@

# Set up output directory

@@

```{r}
## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All") %>%  # to remove summary values from the data
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of historical deaths by year as calculated from different datasets. The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other. The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other. This may be due to the following (verbatim from data description online): Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
```

@@

```{r}
weekly_covid_cases_hb <-
  cases_hb %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    crude_rate_positive_per100k = max(crude_rate_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )

weekly_covid_cases_la <-
  cases_la %>%
  group_by(year, week_number, ref_area) %>%
  summarise(
    weekly_positive = sum(daily_positive),
    cumulative_positive = max(cumulative_positive),
    crude_rate_positive_per100k = max(crude_rate_positive),
    .groups = "drop"
  ) %>%
  mutate(
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)
    )
```

We can use the yearly cumulative case count to group geographic areas by how many cases they've had by the end of 2020.

```{r}
cumulative_case_2020_based_grouping_la <-
  weekly_covid_cases_la %>% 
  filter(ref_area!="Scotland") %>%
  filter(year == 2020) %>% 
  filter(week_number == max(week_number)) %>%  # keep last week
  select(ref_area, cumulative_positive, crude_rate_positive_per100k)

cumulative_case_2020_based_grouping_hb <-
  weekly_covid_cases_hb %>% filter(year == 2020) %>% 
  filter(ref_area!="Scotland") %>%
  filter(year == 2020) %>% 
  filter(week_number == max(week_number)) %>%  # keep last week 
  select(ref_area, cumulative_positive, crude_rate_positive_per100k)

k_means_clustering_cases_and_per100k_2020_hb <- map(
  .x = 2:4,
  .f = ~kmeans(x=cumulative_case_2020_based_grouping_hb %>% select(cumulative_positive, crude_rate_positive_per100k), centers = .x, nstart = 1e3)
) %>% set_names(nm=paste0("k",2:4))

clusters_hb <-
  bind_cols(
    cumulative_case_2020_based_grouping_hb %>% select(ref_area),
    map_dfc(.x = k_means_clustering_cases_and_per100k_2020_hb, .f = function(x) return(x$cluster))
  )

cumulative_case_2020_based_grouping_hb <-
  cumulative_case_2020_based_grouping_hb %>%
  left_join(clusters_hb, by="ref_area") %>%
  mutate(
    quartiles_cases = ntile(cumulative_positive, n=4),
    quartiles_per100k = ntile(crude_rate_positive_per100k, n=4)
  )

# TODO: apply to LA also!

hb_cases_plots <- list()

hb_cases_plots$cumulative_cases_vs_per100k <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
  geom_point() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate"
  ) +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_log10 <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area)) +
  geom_point() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases (log)",
    y="COVID cases per 100,000 population (log)",
    title="Cases vs. population rate (log scale axes)"
  ) +
  scale_x_log10() +
  scale_y_log10() +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_cases))) +
  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
  scale_colour_viridis_d() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate"
  ) +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_quartiles_cases <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(quartiles_per100k))) +
  geom_point(size=3, aes(colour=factor(quartiles_cases))) +
  scale_colour_viridis_d() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate",
    colour = "Quartile (Cases per 100,000)"
  ) +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_k2 <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k2))) +
  geom_point(size=3, aes(colour=factor(k2))) +
  scale_colour_viridis_d() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate"
  ) +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_k3 <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k3))) +
  geom_point(size=3, aes(colour=factor(k3))) +
  scale_colour_viridis_d() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate"
  ) +
  NULL

hb_cases_plots$cumulative_cases_vs_per100k_k4 <-
  cumulative_case_2020_based_grouping_hb %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, shape=factor(k4))) +
  geom_point(size=3, aes(colour=factor(k4))) +
  scale_colour_viridis_d() +
  geom_label_repel() +
  theme_minimal() +
  labs(
    x="COVID cases",
    y="COVID cases per 100,000 population",
    title="Cases vs. population rate"
  ) +
  NULL

hb_cases_plots_assembled <- patchwork::wrap_plots(hb_cases_plots, ncol = 2)

total_deaths_hb %>% filter(year==2020 & str_starts(place_of_death, pattern="Home")) %>%
  left_join(cumulative_case_2020_based_grouping_hb, by="ref_area") %>%
  ggplot(data=., aes(x=cumulative_positive, y=deaths_all_causes, label=ref_area)) +
  geom_point() +
  geom_label_repel() +
  theme_minimal() +
  NULL

# k means clustering
hb_clustered <- kmeans(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland") %>% select(2:3), centers = 4, nstart = 2500)
bind_cols(cumulative_case_2020_based_grouping_hb %>% filter(ref_area!="Scotland"), cluster=hb_clustered$cluster) %>%
  ggplot(data=., aes(x=cumulative_positive, y=crude_rate_positive_per100k, label=ref_area, fill=cluster, colour=cluster)) +
    geom_point() +
    geom_label_repel() +
    theme_minimal() +
    NULL

@@

# Calculate moving averages (rolling mean) over multiple weeks

@@

# Comparisons of annual rates with current rates

```{r}
## this is fairly ad-hoc... but it works!
table_comparing_rates <-
  bind_rows(
    ## all available data
    merged_data_overall %>%
      filter(age == "all" &
               sex == "all" &
               str_starts(place_of_death, "Home")) %>% group_by(year, week_number) %>%
      summarise(
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        .groups = "drop"
      ) %>% left_join(
        past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                         year != "average") %>% group_by(week_number) %>% summarise(
                                           average_weekly_deaths = sum(number_of_deaths) / 5,
                                           .groups = "drop"
                                         ),
        by = "week_number"
      ) %>% summarise(
        period = "all available data",
        deaths_all_causes = sum(deaths_all_causes),
        deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
        deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE),
        past_annual_deaths = sum(average_weekly_deaths)
      )
    ,
    ## 2020 only
    bind_cols(
      merged_data_overall %>%
        filter(
          age == "all" &
            sex == "all" & str_starts(place_of_death, "Home") & year == 2020
        ) %>%
        summarise(
          period = "2020",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(str_starts(place_of_death, "Home") &
                                       year != "average") %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    ),
    ## 2020 from week 12 onwards (including week 12) - pandemic part of 2020
    bind_cols(
      merged_data_overall %>%
        filter(
          age == "all" &
            sex == "all" &
            str_starts(place_of_death, "Home") &
            year == 2020 & week_number >= 12
        ) %>%
        summarise(
          period = "2020, from wk12",
          deaths_all_causes = sum(deaths_all_causes),
          deaths_non_covid = sum(deaths_non_covid, na.rm = TRUE),
          deaths_covid_related = sum(deaths_covid_related, na.rm = TRUE)
        ),
      past_deaths_overall %>% filter(
        week_number >= 12 &
          str_starts(place_of_death, "Home") &
          year != "average"
      ) %>% summarise(past_annual_deaths = sum(number_of_deaths) / 5)
    )
  ) %>%
  mutate(
    ratio_deaths_all_causes_to_historical = deaths_all_causes / past_annual_deaths,
    ratio_non_covid_deaths_to_historical = deaths_non_covid / past_annual_deaths,
    proportion_covid_deaths = deaths_covid_related / deaths_all_causes
  )

write_csv(x = table_comparing_rates, file = file.path(dir_outputs, "annual_death_rates_comparison_with_historical_deaths.csv"))

table_comparing_rates %>% knitr::kable(caption = "Comparing annual home death rate during COVID pandemic to historical deaths at home")
```

@@

@@

```{r}
theme_set(theme_minimal(base_size = 16) +
            theme(panel.grid.minor = element_blank(),
                  axis.title.y = element_text(margin = margin(0, 20, 0, 0)),  # remove gap to the left of y axis title and below the x axis title
                  axis.title.x = element_text(margin = margin(20, 0, 0, 0))
            ))

date_range_text <- glue::glue("Data from w/c {format(min(weekly_deaths_overall$date_w_c), '%d %B %Y')} to w/c {format(max(weekly_deaths_overall$date_w_c), '%d %B %Y')} (inclusive).")

dashed_line_historical_range_text <- "Range of deaths 2015-19 shown as shaded area; mean shown as dashed line."

@@

## All of Scotland, deaths at home

```{r, fig.width=12, fig.height=9, warning=FALSE}
# TODO: add start of lockdowns 1-3?
# TODO: what is that dip at week 38/39? not obvious - start of september was tier 4 restrictions in big LAs
total_deaths_at_home_in_period <- glue::glue("Total deaths at home N={format(merged_deaths_overall %>% filter(sex == 'all' & age == 'all' & place_of_death %in% c('Home & other non-institution')) %>% .$deaths_all_causes %>% sum, big.mark=',')}")

figure_deaths_at_home <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid), 
               names_to = "deaths_type", values_to = "deaths") %>% 
  ggplot(aes(x = week_number_run_over)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  geom_vline(xintercept = 54, linetype = "dotted") +
  scale_x_continuous(breaks = seq(0,60,by=10)) +
  ylim(0, 600) +
  facet_grid(~place_of_death) +
  theme_minimal() +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths", 
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  annotate(geom = "text", label = total_deaths_at_home_in_period, x = 25, y = 575, hjust=0) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(dashed_line_historical_range_text,vertical_line_text, death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       )

## figure for a bit less than half A4 page is approx 14*11cm
walk(
  .x = c(".pdf",".svg",".png"),
  .f = ~ggsave(filename = paste0(dir_outputs,"/figure_deaths_at_home",.x), plot = figure_deaths_at_home, width = 14, height = 11, units = "cm", dpi = 300)
)

@@
